Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-05-25 SIMPaper/sbpaper.tex
cleaned, licensed version of cbnn

\abstract{In survival analysis, covariate interactions are difficult to
model without prior knowledge using regression-based methods. Neural
network-based survival models have been developed for data-driven
interaction modeling, however, these approaches lack implementations for
time-varying interactions or a flexible baseline hazard. To address
this, we extend the case-base sampling framework and propose Case-Base
Neural Networks (CBNN), which introduce time directly into the model,
using commonly available neural network components for a simple
implementation. First, we case-base sample the survival data, converting
the survival time of each participant into discrete moments in time.
This allows us to estimate the probability an event occurs at a given
moment using a neural network to get the hazard. We compare CBNN to
survival methods based on regression and neural networks in two
simulations and two real data applications with a fixed neural network
architecture. We report metrics for each model over follow-up time using
both a proper (Index of Prediction Accuracy) and non-proper (inverse
probability censoring weights-adjusted concordance index) scoring rule.
A simple simulation shows CBNN outperforming the other neural network
methods. The complex simulation highlights the ability of CBNN to model
both a complex baseline hazard and time-varying interactions,
outperforming all competitors. The first real data application shows
CBNN outperforming all neural network competitors, while a second real
data application shows competitive performance. CBNN has shown that it
is more consistent in prediction performance across time in the
simulations and real data applications compared to the competing neural
network approaches.}

@@

The widespread use of the Cox proportional hazards model has led to a
focus on survival models based on hazard ratios and relative risks
rather than on survival curves and absolute risks \citep{hanley2009}. In
clinical settings, Cox models are frequently prioritized over
smooth-in-time accelerated failure time (AFT) models \citep{hanley2009},
which can estimate absolute risks by modeling the hazard directly
through user specification of a baseline hazard distribution. The user
may forgo this choice by implementing a flexible set of parameters using
splines, as proposed by Royston and Palmar \citep{royston2002flexible}.
These regression models also require user-specified interactions, which
may limit risk prediction accuracy.

Over longer follow-up, it is challenging to maintain the proportional
hazards assumption as the risk associated with each covariate may not be
constant with time. Previous studies related to breast cancer have
discovered time-varying interactions with covariates of interest, like
tumor size \citep{coradini2000time} \citep{hilsenbeck1998time}. Another
limitation of regression-based models is the requirement of prior
knowledge; the analyst must know potential time-varying interactions and
explicitly evaluate them.

Data-driven neural network approaches bypass the need for prior
knowledge of interaction terms. DeepSurv is a semi-parametric
proportional hazards neural network model that implements the Cox
partial log-likelihood as a custom loss function
\citep{katzman2018DeepSurv}. This results in a stepwise absolute risk
curve that cannot accommodate time-varying interactions. Compared to Cox
regression, DeepSurv has shown a small improvement in performance using
a single layer neural network on the Study to Understand Prognoses
Preferences and Risk Treatments (SUPPORT) dataset
\citep{knaus1995SUPPORT}. Farraggi et al.~suggested a modification to
the loss function required for a Cox neural network that can model
non-proportional hazards \citep{faraggi1995neural}. Wang et al.~took the
concept of extreme learning machines and applied them in a survival
context, removing the need for tuning the number of layers and nodes
\citep{wang2019extreme}. A Cox-based neural network combined with Local
Interpretable Model-Agnostic Explanations (LIME)
\citep{ribeiro2016model} interprets the association of each Single
Nucleotide Polymorphism (SNP) \citep{sun2020genome} to a phenotype.

DeepHit requires each survival time of interest to be specified in the
model and directly estimates survival curves, rather than deriving a
hazard function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and its concordance index (c-index)
score outperforms DeepSurv \citep{katzman2018DeepSurv} on the Molecular
Taxonomy of Breast Cancer International Consortium (METABRIC) dataset
\citep{curtis2012genomic}.

Ripley et al.~presented seven different distributions and modeling
paradigms based on neural networks \citep{ripley2004non}. They provided
the methodology behind parametric log-logistic and log-normal neural
network models. Like DeepHit, these models are restricted to one of the
seven baseline hazard distributions, reducing the flexibility of the
average survival prediction.

Deep Survival Machines (DSM) provides a flexible, parametric survival
model using neural networks with a mixture of distributions as the
baseline hazard \citep{dsmPaper}. DSM allows the user to specify a
mixture of distributions for a flexible baseline hazard. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

These methods need custom loss and activation functions that are
difficult to implement in practice. We note the complexity of these
alternative neural network approaches; DeepSurv, DeepHit, and DSM all
require custom loss functions \citep{katzman2018DeepSurv}
\citep{lee2018DeepHit} \citep{dsmPaper}, while DSM requires user
implementations for distributions beyond Log-Normal or Weibull and a
two-phase learning process \citep{dsmPaper}. Regression-based approaches
require prior specification of all interaction terms, which makes it
challenging to model covariate effects that change over time. In this
article, we propose Case-Base Neural Networks (CBNN) as a method that
models time-varying interactions and a flexible baseline hazard using
commonly available neural network components. Our approach to modeling
the full hazard uses case-base sampling \citep{hanley2009}, which allows
probabilistic models to predict survival outcomes. Following the
case-base framework, we can use transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by
including splines of time as covariates, we can approximate the
Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}. An alternative approach to Royston-Parmar
is case-base with logistic regression (CBLR), which provides some
flexibility in modeling, but still requires user specification of
time-varying interactions and baseline hazard, while CBNN can model both
without user specification; only a sufficiently deep network is
required.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choice and software implementation. Then, in Section
\ref{sims}, we describe our metrics and compare the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression, and case-base using logistic
regression on simulated data. Next, Section \ref{casestudies} describes
the real-data analysis, while in Section \ref{discussion}, we explore
the implications of our results and contextualize them within neural
network survival analysis in a single event setting.

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics, and
software}\label{methods}}

@@

For each person-moment sampled, let \(X_i\) be the corresponding
covariate profile \(\left(x_{i1},x_{i2},...,x_{ip} \right)\), \(T_i\) be
the time of the person-moment, and \(Y_i\) be the indicator variable for
whether the event of interest occurred at time \(T_i\). We estimate the
hazard function \(h(t \mid X_i)\) using the sampled person-moments.
Recall that \(h(t \mid X_i)\) is the instantaneous potential of
experiencing the event at time \(t\) for a given set of covariates
\(X_i\), assuming \(T_i \geq t\).

Now, let \(b\) be the (user-defined) size of the \emph{base series}, and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly, then the hazard
function of the sampling process is equal to \(b/B\). Therefore, we have
the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

We may estimate the hazard function if we adjust for the bias introduced
when sampling a fraction of the study base \(B\) by adding the offset
term \(\log\left(\frac{b}{B} \right)\) (\eqref{eqn:offset}). Next, we
propose using neural networks to model the odds.

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By allowing a neural network
to approximate a higher-order polynomial of time, the baseline hazard
specification is now data-driven, where user-defined hyperparameters
such as regularization, number of nodes, and layers control the
flexibility of the model. We provide a detailed description of the
choices we made in the next sub-section.

@@

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/pmnnProjects/paper/figures/nnarch2}

@@

The following derivation shows how our probability estimate is converted
to odds: \begin{align*}
 \log\left( h(t \mid X_i) \right) &= \log\left(\frac{\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function \citep{gulli2017}:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*} where \(\hat{f}_{\theta}(x_{i}, t_{i})\) is our estimate
for a given covariate profile and time, \(y_{i}\) is our target value
specifying whether an event occurred, and \(N\) represents the number of
individuals in our training set.

Backpropagation is used to optimize the parameters in the model with an
appropriate minimization algorithm (e.g.~Adam, RMSPropagation,
stochastic gradient descent)\citep{gulli2017}. For our analysis, we use
Adam. Note that the size of the \emph{case series} is fixed as the
number of events, but we can make the \emph{base series} as large as we
want. A ratio of 100:1 \emph{base series} to \emph{case series} is
sufficient \citep{hanley2009}. We pass our feed-forward neural network
through a sigmoid activation function (Figure \ref{fig:NNarch}).
Finally, we can convert this model output to a hazard. When using our
model for predictions, we manually set the offset term to 0 in the new
data, as we account for the bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the
risk function (\(F\)) at time \(t\) for a covariate profile \(X\),
viz.\begin{align}\label{eqn:ci2}
F\left(t|X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} We use Riemann's Sum \citep{hughes2020calculus} to
approximate the integral in \eqref{eqn:ci2}.

@@

Neural networks provide flexibility when defining the architecture and
optimization parameters of our models. These hyperparameter choices can
affect the estimated parameters and were chosen during a set of initial
simulations to determine if CBNN can learn complex interactions in
practice. We set a max number of epochs to be \(2000\), batch size as
\(512\), learning rate as \(10^{-3}\), decay as \(10^{-7}\), patience as
\(10\) epochs, \(\{50,50,25,25\}\) nodes in each hidden layer with
\(50\%\) dropout at each layer, a validation split of \(20\%\), a
testing split of \(10%
\), a minimum delta loss on the validation set of \(10^{-7}\) over 10
epochs, and Adam \citep{gulli2017} as our optimizer. These choices may
permit the model to approximate higher-order interactions while
preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test). We select
the best weights after training based on validation loss
\citep{gulli2017}.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper} and the DeepHit model
using \textbf{pycox} \citep{lee2018DeepHit}. We made the components of
CBNN using the \textbf{keras} \citep{keras} package and the
\textbf{casebase} package for the sampling step. The \textbf{simsurv}
package \citep{simsurv} is used to simulate our simulation studies,
while \textbf{flexsurv} \citep{flexsurv} is used to fit a flexible
baseline hazard using splines for our complex simulation. We use the
implementation of c-ipcw from the python package \textbf{sksurv}
\citep{sksurv}. The \textbf{riskRegression} package
\citep{riskRegression} is used to get the Brier score and IPA metrics.
We changed the \textbf{riskRegression} package to be used with any user
supplied risk function \(F\). To ensure that both R and Python-based
models are running in unison on the same data through our simulations
and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

In this section, we use simulated data to evaluate the performance of
CBNN and compare our approach with existing methods based on regression
(Cox, CBLR) and neural network (DeepHit, DeepSurv, DSM) methods. We
specify a linear combination of each covariate as the linear predictor
in regression-based approaches (Cox, CBLR), which contrasts with neural
network approaches that allow for non-linear interactions. We simulate
data under a simple exponential model, and a complex baseline hazard
with time-varying interactions, each with 10\% random censoring. For
both settings, we simulate three covariates:

@@

The DeepHit-specific hyperparameter alpha is set to \(0.5\) (equal
weight between its negative log-likelihood and ranking losses
\citep{lee2018DeepHit}). We change the \textbf{DeepSurvivalMachines}
\citep{dsmPaper} package to include dropout and a minimum delta loss
during the fitting process. For DSM, we define a mixture model of six
Weibull distributions for the baseline hazard. All other hyperparameters
are held constant across all neural network methods in both the
simulation studies and real data applications.

Besides the methods mentioned above, we include the Optimal model in our
comparisons using CBLR. That is, we include the exact functional form of
the covariates in a CBLR model (referred to as Optimal for simplicity).
We calculate \(t\)-based 95\% confidence intervals using 100
replications of the simulated data For all analyses, we use 80\% for
training and 20\% for the test set. 20\% of the training set is kept for
validation at each epoch. We predict risk functions \(F\) using
\eqref{eqn:ci2} for individuals in the test set, which are used to
calculate our c-ipcw and IPA scores.

@@

We use two metrics to assess the performance of the different methods of
interest: 1) index of prediction accuracy (IPA) \citep{kattan2018index}
and 2) inverse probability censoring weights-adjusted concordance index
(c-ipcw) \citep{uno2011}, which we define below. Both these
time-dependent metrics provide transparency as to when in follow-up time
each model may perform better than the others.

@@

The IPA is a function of the Brier score (\(BS(t)\)) \citep{graf1999},
which is defined as \begin{align}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(\widehat{F}(X_{i},t)-1\right)^{2}\cdot I_{T_{i}\leq t,\delta_{i}=1}}{\widehat{G}(T_{i})} + \frac{\left(\widehat{F}(X_{i},t)\right)^{2}\cdot I_{T_{i}>t}}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have
experienced the event, \(N\) represents the number of samples in our
dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[C>t]\) is
a non-parametric estimate of the censoring distribution, \(C\) is
censoring time and \(T_{i}\) is an individual's survival or censoring
time. The Brier score provides a score that accounts for the information
loss because of censoring. There are three categories of individuals
that may appear within the dataset once we fix our \(t\) of interest.
Individuals who experienced the event before \(t\) are present in the
first term of the equation. The second term of the equation includes
individuals who experience the event or are censored after \(t\). Those
censored before \(t\) are the third category of people. The inverse
probability censoring weights (IPCW) adjustment (\(G(\cdot)\)) is to
account for these category three individuals whose information is
missing. The IPA score as a function of time is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the Brier score over time
\(t\) for the model of interest, and \(BS_{null}(t)\) represents the
Brier score if we use an unadjusted Kaplan-Meier curve as the prediction
for all observations \citep{kattan2018index}. Note that IPA has an upper
bound of one, where positive values show an increase in performance over
the null model and negative values show that the null model performs
better. These scores demonstrate how performance changes over follow-up
time.

A potential artifact of IPA is that the score is unstable at earlier and
later survival times. This is because of a near equivalent Brier score
between each model and the null model. At small values, a difference of
\(0.1\) creates a more significant fold change than at larger values. As
the Brier score is potentially small at the start and at the end of
their respective curves, we may see instability of the IPA score at
those ends.

>>>>

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While this flexibility has led
to methods with better predictive performance than regression-based
approaches, current neural networks cannot model both time-varying
interactions and complex baseline hazards. To address this, we propose
Case-Base Neural Networks (CBNN) as a new approach that combines the
case-base sampling framework with flexible architectures. Our method
naturally accounts for censoring, does not require additional
hyperparameters and uses binary cross-entropy as the loss function.
Using a novel sampling scheme and data augmentation, we incorporate time
directly into a feed-forward neural network. CBNN predicts the
probability of an event occurring at a given moment and estimates the
hazard function. We compare the performance of CBNN to survival methods
based on regression and neural networks in two simulations and two real
data applications. We report two time-dependent metrics for each model.
In the simulations and real data applications, CBNN provides a more
consistent predictive performance across time and outperforms the
competing neural network approaches. For a simple simulation with an
exponential hazard model, CBNN consistently outperforms the other neural
network methods. For a complex simulation, which highlights the ability
of CBNN to model both a complex baseline hazard and time-varying
interactions, CBNN outperforms all competitors. The first real data
application shows CBNN outperforming all neural network competitors,
while a second real data application shows competitive performance. This
study highlights the benefit of combining case-base sampling with deep
learning to provide a simple and flexible modeling framework for
data-driven, time-varying interaction modeling of survival outcomes.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate
absolute risks by modeling the hazard directly through user-specified
baseline hazard distribution \citep{kleinbaum2012survival}. Cox
proportional hazards models are used more often than AFT models, causing
analyses to be based on hazard ratios and relative risks rather than on
survival curves and absolute risks \citep{hanley2009}. The
identification of an appropriate distribution for the baseline hazard in
an AFT model may be difficult for common diseases that have many
interacting risk factors or a Cox model whose pathogenesis may change
with age, making it difficult to maintain the proportional hazards
assumption. For example, previous studies of breast cancer incidence
have discovered time-varying interactions with covariates of interest,
including tumor size \citep{coradini2000time}. One approach to resolve
this involves using the basis of splines on time to provide flexibility
in the baseline hazard \citep{royston2002flexible}. However,
regression-based models are limited in that they require prior knowledge
about potential time-varying interactions and their quantitative
effects.

Neural networks provide a data-driven approach to approximating
interaction terms. For example, DeepSurv is a neural network-based
proportional hazards model that implements the Cox partial
log-likelihood as a custom loss function \citep{katzman2018DeepSurv},
resulting in a stepwise absolute risk curve that cannot accommodate
time-varying interactions. Compared to Cox regression, DeepSurv shows
better performance on the Study to Understand Prognoses Preferences and
Risk Treatments (SUPPORT) dataset \citep{knaus1995SUPPORT}. To handle
non-proportional hazards, a modification to the loss function was
proposed \citep{faraggi1995neural}. To remove the need for tuning the
number of layers and nodes, the concept of extreme learning machines has
been applied to a Cox neural network model \citep{wang2019extreme}.

As an alternative method that assumes a baseline hazard distribution,
DeepHit specifies each survival time of interest in the model and
directly estimates survival curves, rather than deriving a hazard
function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and it outperforms DeepSurv on the
Molecular Taxonomy of Breast Cancer International Consortium (METABRIC)
dataset \citep{curtis2012genomic}.

Providing further flexibility, Deep Survival Machines (DSM) is a
parametric survival model using neural networks with a mixture of
distributions as the baseline hazard \citep{dsmPaper}. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

We note that these alternative neural network approaches all require
custom loss functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
\citep{dsmPaper}. DeepHit introduces a new hyperparameter weighing its
two loss functions (negative log-likelihood and ranking losses) while
DSM requires a two-phase learning process and user implementations for
distributions beyond Log-Normal or Weibull \citep{lee2018DeepHit}
\citep{dsmPaper}. Regression-based approaches require prior
specification of all interaction terms, which makes it challenging to
model covariate effects that change over time. The current neural
network models provide flexibility at the cost of opacity, while
regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method
that models time-varying interactions and a flexible baseline hazard
using commonly available neural network components. Our approach to
modeling the full hazard uses case-base sampling \citep{hanley2009}.
This sampling technique allows probabilistic models to predict survival
outcomes. As part of the case-base framework, we use transformations of
time as a feature (covariate) to specify different baseline hazards. For
example, by including splines of time as covariates, we can approximate
the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}, however, this still requires explicit use
of time-varying interactions. CBNN can model both without user
specification; only a sufficiently deep network is required.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choices and software implementation. Section
\ref{sims}, describes our metrics and compares the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression and case-base using logistic
regression (CBLR) on simulated data. Section \ref{casestudies} describes
the real-data analysis, while Section \ref{discussion} explores the
implications of our results and contextualizes them within neural
network survival analysis in a single event setting.

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics and
software}\label{methods}}

@@

For each person-moment sampled, let \(X_i\) be the corresponding
covariate profile \(\left(x_{i1},x_{i2},...,x_{ip} \right)\), \(T_i\) be
the time of the person-moment and \(Y_i\) be the indicator variable for
whether the event of interest occurred at time \(T_i\). We estimate the
hazard function \(h(t \mid X_i)\) using the sampled person-moments.
Recall that \(h(t \mid X_i)\) is the instantaneous potential of
experiencing the event at time \(t\) for a given set of covariates
\(X_i\), assuming \(T_i \geq t\).

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjusting for the bias
introduced when sampling a fraction of the study base \(B\) by adding
the offset term \(\log\left(\frac{b}{B} \right)\) as in
\eqref{eqn:offset}. Next, we propose using neural networks to model the
odds.

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the model. We provide a detailed description
of the choices we made in the next sub-section.

@@

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/cbnnManuscript/figures/nnarch2}

@@

The following derivation shows how our probability estimate is converted
to odds: \begin{align*}
 \log\left( h(t \mid X) \right) &= \log\left(\frac{\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function \citep{gulli2017}:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*} where \(\hat{f}_{\theta}(x_{i}, t_{i})\) is our estimate
for a given covariate profile and time, \(y_{i}\) is our target value
specifying whether an event occurred and \(N\) represents the number of
individuals in our training set.

Backpropagation with an appropriate minimization algorithm (e.g.~Adam,
RMSPropagation, stochastic gradient descent) is used to optimize the
parameters in the model \citep{gulli2017}. For our analysis, we use Adam
as implemented in Keras \citep{gulli2017}. Note that the size of the
\emph{case series} is fixed as the number of events, but we can make the
\emph{base series} as large as we want. A ratio of 100:1 \emph{base
series} to \emph{case series} is sufficient \citep{hanley2009}. We pass
our feed-forward neural network through a sigmoid activation function
(Figure \ref{fig:NNarch}). Finally, we can convert this model output to
a hazard. When using our model for predictions, we manually set the
offset term to 0 in the new data, as we account for the bias during the
fitting process.

Since we are directly modeling the hazard, we can readily estimate the
risk function (\(F\)) at time \(t\) for a covariate profile \(X\),
viz.\begin{align}\label{eqn:ci2}
F\left(t\mid X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} We use a finite Riemann sum \citep{hughes2020calculus} to
approximate the integral in \eqref{eqn:ci2}.

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter decisions can affect the
estimated parameters and were chosen during a set of initial simulations
to determine if CBNN can learn complex interactions in practice. We set
a max number of epochs to be \(2000\), batch size as \(512\), learning
rate as \(10^{-3}\), decay as \(10^{-7}\), patience as \(10\) epochs,
\(\{50,50,25,25\}\) nodes in each hidden layer with \(50\%\) dropout at
each layer, a validation split of \(20\%\), a testing split of \(10%
\), a minimum delta loss on the validation set of \(10^{-7}\) over 10
epochs and Adam \citep{gulli2017} as our optimizer. These choices may
permit the model to approximate higher-order interactions while
preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test) for each
method. We select the best weights after training based on validation
loss \citep{gulli2017}.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DeepHit model using \textbf{pycox}
\citep{lee2018DeepHit} and the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. We use the implementation of \(C_{IPCW}\)??? from the python
package \textbf{sksurv} \citep{sksurv}. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric). Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

In this section, we use simulated data to evaluate the performance of
CBNN and compare our approach with existing regression-based (Cox, CBLR)
and neural network-based (DeepHit, DeepSurv, DSM) methods. We specify a
linear combination of each covariate as the linear predictor in
regression-based approaches (Cox, CBLR), which contrasts with neural
network approaches that allow for non-linear interactions. We simulate
data under a simple exponential model and a complex baseline hazard with
time-varying interactions, each with 10\% random censoring. For both
settings, we simulate three covariates:

@@

The DeepHit-specific hyperparameter alpha is set to \(0.5\) (equal
weight between its negative log-likelihood and ranking losses
\citep{lee2018DeepHit}). We modify the \textbf{DeepSurvivalMachines}
\citep{dsmPaper} package to include dropout and a minimum delta loss
during the fitting process. For DSM, we define a mixture model of six
Weibull distributions for the baseline hazard. All other hyperparameters
are held constant across all neural network methods in both the
simulation studies and real data applications.

Besides the methods mentioned above, we include the Optimal model in our
comparisons using CBLR. That is, we include the exact functional form of
the covariates in a CBLR model (referred to as Optimal for simplicity).
We calculate \(t\)-based 95\% confidence intervals using 100
replications of the simulated data. For all analyses, we use 80\% for
training and 20\% for the test set. 20\% of the training set is kept for
validation at each epoch. We predict risk functions \(F\) using
\eqref{eqn:ci2} for individuals in the test set, which are used to
calculate our \(C_{IPCW}\) and IPA scores.

@@

We use two metrics to assess the performance of the different methods of
interest: 1) (IPA) \citep{kattan2018index} and 2) inverse probability
censoring weights-adjusted concordance index (\(C_{IPCW}\))
\citep{uno2011}, which we define below. Both these time-dependent
metrics provide transparency as to when in follow-up time each model may
perform better than the others.

@@

The IPA is a function of the Brier score (\(BS(t)\)) \citep{graf1999},
which is defined as \begin{align}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have
experienced the event, \(N\) represents the number of samples in our
dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[c>t]\) is
a non-parametric estimate of the censoring distribution, \(c\) is
censoring time and \(T_{i}\) is an individual's survival or censoring
time. The Brier score provides a score that accounts for the information
loss because of censoring. There are three categories of individuals
that may appear within the dataset once we fix our \(t\) of interest.
Individuals who experienced the event before \(t\) are present in the
first term of the equation. The second term of the equation includes
individuals who experience the event or are censored after \(t\). Those
censored before \(t\) are the third category of people. The inverse
probability censoring weights (IPCW) adjustment (\(G(\cdot)\)) is to
account for these category three individuals whose information is
missing. The IPA score as a function of time is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the Brier score over time
\(t\) for the model of interest and \(BS_{null}(t)\) represents the
Brier score if we use an unadjusted Kaplan-Meier (KM) curve as the
prediction for all observations \citep{kattan2018index}. Note that IPA
has an upper bound of one, where positive values show an increase in
performance over the null model and negative values show that the null
model performs better. These scores show how performance changes over
follow-up time.

A potential artifact of IPA is that the score is unstable at earlier and
later survival times. This is because of a near equivalent Brier score
among each model and the null model. At small values, a difference of
\(0.1\) creates a more significant fold change than at larger values. As
the Brier score is potentially small at the start and at the end of
their respective curves, The IPA score may be unstable at the same
locations.
